pdp: An R Package for Constructing Partial Dependence Plots

### 2016-12-29 manuscript/greenwell.tex
minor edits

\citet{harrison-1978-hedonic} were among the first to analyze the well-known Boston housing data. One of their goals was to find a housing value equation using data on median housing values from $n = 506$ census tracts in the suburbs of Boston from the 1970 census; see Table IV of \citet{harrison-1978-hedonic} for a description of each variable. The data violate many classical assumptions like linearity, normality, and constant variance. Nonetheless, \citeauthor{harrison-1978-hedonic}---using a combination of transformations, significance testing, and grid searches---were able to find a reasonable fitting model ($R^2 = 0.81$). Part of the payoff for there efforts was an interpretable prediction equation which is reproduced in Equation~\eqref{eqn:boston}.
\begin{equation}
\label{eqn:boston}
\begin{split}
\widehat{\log\left(MV\right)} &= 9.76 + 0.0063 RM^2 + 8.98\times10^{-5} AGE - 0.19\log\left(DIS\right) + 0.096\log\left(RAD\right) \\
  & \quad - 4.20\times10^{-4} TAX - 0.031 PTRATIO + 0.36\left(B - 0.63\right)^2 - 0.37\log\left(LSTAT\right) \\
  & \quad - 0.012 CRIM + 8.03\times10^{-5} ZN + 2.41\times10^{-4} INDUS + 0.088 CHAS - 0.0064 NOX^2
\end{split}
\end{equation}

@@

\begin{algorithm}
\begin{enumerate}
  \item For $i \in \left\{1, 2, \dots, k\right\}$:
  \begin{enumerate}
    \item Copy the training data and replace the original values of $x_1$ with the constant $x_{1i}$.
    \item Compute the vector of predicted values from the modified copy of the training data.
    \item Compute the average prediction to obtain $\bar{f}_1\left(x_{1i}\right)$.
  \end{enumerate}
  \item Plot the pairs $\left\{x_{1i}, \bar{f}_1\left(x_{1i}\right)\right\}$ for $i = 1, 2, \dotsc, k$.
\end{enumerate}
\caption{A simple algorithm for consructing the partial dependence of the response on a single predictor $x_1$. \label{alg:pdp}}
\end{algorithm}
Algorithm~\ref{alg:pdp} can be quite computationally intensive since it involves $k$ passes over the training records. Fortunately, this can be parallelized quite easily (more on this in a later section). This algorithm can also be easily extended to larger subsets of two or more features as well.

@@

The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for various types of fitted models (not just random forests), (2) it will allow for any number of predictors to be used (e.g., multivariate displays), and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section. For example, the following code chunk uses the random forest model to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv}. The \code{grid.arrange} function is then used to display three PDPs, which make use of various \code{plotPartial} options\footnote{See Section~\ref{sec:computational} for an example of how to add a label to the colorkey in these types of graphs.}, on the same graph. The result is displayed in Figure~\ref{fig:pd_lstat_rm}.
\begin{example}
rwb <- colorRampPalette(c("red", "white", "blue"))
pd <- partial(boston.rf, pred.var = c("lstat", "rm"))
grid.arrange(
  # Figure 3 (left)
  plotPartial(pd),
  # Figure 3 (middle)
  plotPartial(pd, contour = TRUE, col.regions = rwb),
  # Figure 3 (right)
  plotPartial(pd, levelplot = FALSE, zlab = "cmedv", drape = TRUE,
              colorkey = TRUE, screen = list(z = -20, x = -60)),
  ncol = 3
)
\end{example}
\strong{Note:} the default color map for level plots is the color blind-friendly matplotlib \citep{hunter-matplotlib-2007} 'viridis' color map provided by the \CRANpkg{viridis} package \citep{viridis-pkg}.

@@

Constructing PDPs can be quite computationally expensive. Several strategies are available to ease the computational burden in larger problems. For example, there is no need to compute partial dependence of \code{cmedv} using each unique value of \code{rm} in the training data (which would require $k = 446$ passes over the data!). We could get very reasonable results using a reduced number of points. Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points can be controlled using the \code{grid.resolution} option in the call to \code{partial}. Alternatively, a user-specified grid of values (e.g., containing specific quantiles of interest) can be supplied through the \code{pred.grid} argument. To demonstrate, the following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} using each option, then uses \code{grid.arrange} to display all three PDPs on the same page; the results are displayed in Figure~\ref{fig:partial_manual}.
\begin{example}
grid.arrange(
  # Figure 5 (left)
  partial(boston.rf, "rm", plot = TRUE),
  # Figure 5 (middle)
  partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE),
  # Figure 5 (right)
  partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE),
  ncol = 3
)
\end{example}

@@

Using the \code{pred.fun} argument it is possible to obtain PDPs for classification problems on the probability scale; we just need to write a function that extracts the predicted class probabilities of interest and then averages the result. The snippet of code below uses the fitted SVM from the iris example of Section~\ref{sec:class} to extract the average predicted probability of belonging to the \code{Setosa} class.
\begin{example}
pred.prob <- function(object, newdata) {  # see ?predict.svm
  pred <- predict(object, newdata, probability = TRUE)
  prob.setosa <- attr(pred, which = "probabilities")[, "setosa"]
  mean(prob.setosa)
}
\end{example}
Next, we simply pass this function via the \code{pred.fun} argument in the call to \code{partial}. The following chunk of code uses \code{partial} and \code{grid.arrange} to construct PDPs for \code{Petal.Width} and \code{Petal.Length} on the probability scale. The results are displayed in Figure~\ref{fig:partial_iris_prob}.
\begin{example}
grid.arrange(
  # Figure 8 (left)
  partial(iris.svm, pred.var = "Petal.Width", pred.fun = pred.prob, plot = TRUE),
  # Figure 8 (middle)
  partial(iris.svm, pred.var = "Petal.Length", pred.fun = pred.prob, plot = TRUE),
  # Figure 8 (right)
  partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
          pred.fun = pred.prob, plot = TRUE),
  ncol = 3
)
\end{example}

@@

\newpage

For regression problems, the default prediction function is essentially
\begin{example}
pred.fun <- function(object, newdata) {
  mean(predict(object, newdata), na.rm = TRUE)
}
\end{example}
This corresponds to step step 1. (c) in Algorithm~\ref{alg:pdp}. Suppose we would like ICE curves instead. To accomplish this we simply need to pass a prediction function that returns a vector of predictions, one for each observation in \code{newdata}. This is demonstarted for the Boston housing example in the code snippet below.
\begin{example}
pred.ice <- function(object, newdata) {  # all predictions
  predict(object, newdata)
}
lstat.ice <- partial(boston.randomForest, pred.var = "lstat",
                     pred.fun = pred.ice)  # Figure 9 (left)
\end{example}

When the function supplied to \code{pred.fun} returns multiple predictions, the data frame returned by \code{partial} includes an additional column, \code{yhat.id}, that indicates which observation a curve belongs to; in this example, there is one curve for each unique observation in \code{boston}. To further illustrate, the code below constructs ICE curves for specific quantiles of the predictions. The results, along with the traditional PDP for \code{lstat}, are displayed in Figure~\ref{fig:partial_boston_ice}.
\begin{example}
pred.ice.quan <- function(object, newdata) {  # quantiles of predictions
  quantile(predict(object, newdata), probs = 1:9/10)
}
lstat.ice.quan <- partial(boston.randomForest, pred.var = "lstat",
                          pred.fun = pred.ice.quan)  # Figure 9 (middle)
\end{example}

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{partial_boston_ice}
  \caption{ICE curves and PDP depicting the relationship between \code{cmedv} and \code{lstat} for the Boston housing example (the same $y$-axis limits are used for each plot). \textit{Left}: ICE curves. \textit{Middle}: Specific quantiles of ICE curves. \textit{Right}: Standard PDP.}
  \label{fig:partial_boston_ice}
\end{figure}

Using \CRANpkg{dplyr} \citep{dplyr-pkg}, it is rather straightforward to post-process ICE curves to obtain c-ICE curves; see Equation (4) on page 49 of \citet{goldstein-peeking-2015}. The code snippet below illustrates this for the Boston housing example using the predictor \code{age}.
\begin{example}
# Use partial to obtain ICE curves
pred.ice <- function(object, newdata) predict(object, newdata)
age.ice <- partial(boston.randomForest, pred.var = "age", pred.fun = pred.ice)

# Post-process age.ice to obtain c-ICE curves
library(dplyr)
age.ice <- age.ice %>%
  group_by(yhat.id) %>%
  mutate(yhat.centered = yhat - first(yhat))
\end{example}
Furthermore, using \pkg{ggplot2}, we can easily display ICE curves along with the PDP. This can be accomplished by using the \code{stat\_summary} function to average the ICE curves together. The code snippet below uses \pkg{ggplot2} and \code{grid.arrange} to plot ICE curves and c-ICE curves for \code{age} in the Boston housing example. The results are displayed in Figure~\ref{fig:partial_boston_ice_pdp}; compare these to Figures 3-4 in \citet{goldstein-peeking-2015}.
\begin{example}
# ICE curves
p1 <- ggplot(age.ice, aes(age, yhat)) +
  geom_line(aes(group = yhat.id), alpha = 0.3) +
  stat_summary(fun.y = mean, geom = "line", col = "red", size = 2)

# c-ICE curves
p2 <- ggplot(age.ice, aes(age, yhat.centered)) +
  geom_line(aes(group = yhat.id), alpha = 0.3) +
  stat_summary(fun.y = mean, geom = "line", col = "red", size = 2)

# Figure 10
grid.arrange(p1, p2, ncol = 2)
\end{example}
\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{partial_boston_ice_pdp}
  \caption{ICE curves (black curves) and PDP (red curve) depicting the relationship between \code{cmedv} and \code{age} for the Boston housing example. \textit{Left}: Uncentered. \textit{Right}: Centered.}
  \label{fig:partial_boston_ice_pdp}
\end{figure}

@@

# Train an XGBoost model
set.seed(202)
boston.xgb <- train(x = data.matrix(subset(boston, select = -cmedv)),
                    y = boston$cmedv,
                    method = "xgbTree",
                    metric = "RMSE",
                    trControl = ctrl,
                    tuneGrid = xgb.grid)
\end{example}
The model with the smallest root-mean-squared error (RMSE) had \code{nrounds = 3000}, \code{max\_depth = 3}, \code{eta = 0.01}, and \code{subsample = 0.75}; the final model achieved a cross-validated RMSE and $R^2$ of 2.789 and 0.906, respectively. The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together, and then uses \code{grid.arrange} to display all three PDPs on the same page. The results are displayed in Figure~\ref{fig:boston_xgb}.
\begin{example}
grid.arrange(
  # Figure 8 (left)
  partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE),
  # Figure 8 (middle)
  partial(boston.xgb, pred.var = "lstat", plot = TRUE, rug = TRUE),
  # Figure 8 (right)
  partial(boston.xgb, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE),
  ncol = 3
)
\end{example}

>>>>

\citet{harrison-1978-hedonic} were among the first to analyze the well-known Boston housing data. One of their goals was to find a housing value equation using data on median housing values from $n = 506$ census tracts in the suburbs of Boston from the 1970 census; see Table IV of \citet{harrison-1978-hedonic} for a description of each variable. The data violate many classical assumptions like linearity, normality, and constant variance. Nonetheless, \citeauthor{harrison-1978-hedonic}---using a combination of transformations, significance testing, and grid searches---were able to find a reasonable fitting model ($R^2 = 0.81$). Part of the payoff for there time and efforts was an interpretable prediction equation which is reproduced in Equation~\eqref{eqn:boston}.
\begin{equation}
\label{eqn:boston}
\begin{aligned}
\widehat{\log\left(MV\right)} &= 9.76 + 0.0063 RM^2 + 8.98\times10^{-5} AGE - 0.19\log\left(DIS\right) + 0.096\log\left(RAD\right) \\
  & \quad - 4.20\times10^{-4} TAX - 0.031 PTRATIO + 0.36\left(B - 0.63\right)^2 - 0.37\log\left(LSTAT\right) \\
  & \quad - 0.012 CRIM + 8.03\times10^{-5} ZN + 2.41\times10^{-4} INDUS + 0.088 CHAS \\
  & \quad - 0.0064 NOX^2
\end{aligned}
\end{equation}

@@

\begin{algorithm}
\begin{enumerate}
  \item For $i \in \left\{1, 2, \dots, k\right\}$:
  \begin{enumerate}
    \item Copy the training data and replace the original values of $x_1$ with the constant $x_{1i}$.
    \item Compute the vector of predicted values from the modified copy of the training data.
    \item Compute the average prediction to obtain $\bar{f}_1\left(x_{1i}\right)$.
  \end{enumerate}
  \item Plot the pairs $\left\{x_{1i}, \bar{f}_1\left(x_{1i}\right)\right\}$ for $i = 1, 2, \dotsc, k$.
\end{enumerate}
\caption{A simple algorithm for consructing the partial dependence of the response on a single predictor $x_1$. \label{alg:pdp}}
\end{algorithm}
Algorithm~\ref{alg:pdp} can be quite computationally intensive since it involves $k$ passes over the training records. Fortunately, this can be parallelized quite easily (more on this in Section~\ref{sec:computational}). This algorithm can also be easily extended to larger subsets of two or more features as well.

@@

The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for various types of fitted models (not just random forests), (2) it will allow for any number of predictors to be used (e.g., multivariate displays), and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section. For example, the following code chunk uses the random forest model to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv}. The \code{grid.arrange} function is used to display three PDPs, which make use of various \code{plotPartial} options\footnote{See Section~\ref{sec:computational} for an example of how to add a label to the colorkey in these types of graphs.}, on the same graph. The results are displayed in Figure~\ref{fig:pd_lstat_rm}.
\begin{example}
# Compute partial dependence data for lstat and rm
pd <- partial(boston.rf, pred.var = c("lstat", "rm"))

# Default PDP
pdp1 <- plotPartial(pd)

# Add contour lines and use a different color palette
rwb <- colorRampPalette(c("red", "white", "blue"))
pdp2 <- plotPartial(pd, contour = TRUE, col.regions = rwb)

# 3-D surface
pdp3 <- plotPartial(pd, levelplot = FALSE, zlab = "cmedv", drape = TRUE,
                    colorkey = TRUE, screen = list(z = -20, x = -60))

# Display plots side by side (Figure 3)
grid.arrange(pdp1, pdp2, pdp3, ncol = 3)
\end{example}
\strong{Note:} the default color map for level plots is the color blind-friendly matplotlib \citep{hunter-matplotlib-2007} 'viridis' color map provided by the \CRANpkg{viridis} package \citep{viridis-pkg}.

@@

Constructing PDPs can be quite computationally expensive\footnote{The exception is regression trees based on single-variable splits which can make use of the efficient weighted tree traversal method described in \citet{friedman-2001-greedy}, however, only the \pkg{gbm} package seems to make use of this approach; consequently, \pkg{pdp} can also exploit this strategy when used with \pkg{gbm} models (see \code{?partial} for details).} Several strategies are available to ease the computational burden in larger problems. For example, there is no need to compute partial dependence of \code{cmedv} using each unique value of \code{rm} in the training data (which would require $k = 446$ passes over the data!). We could get very reasonable results using a reduced number of points. Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points can be controlled using the \code{grid.resolution} option in the call to \code{partial}. Alternatively, a user-specified grid of values (e.g., containing specific quantiles of interest) can be supplied through the \code{pred.grid} argument. To demonstrate, the following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} using each option, then uses \code{grid.arrange} to display all three PDPs on the same page; the results are displayed in Figure~\ref{fig:partial_manual}.
\begin{example}
# Display plots side by side (Figure 5)
grid.arrange(
  partial(boston.rf, "rm", plot = TRUE),
  partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE),
  partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE),
  ncol = 3
)
\end{example}

@@

Using the \code{pred.fun} argument it is possible to obtain PDPs for classification problems on the probability scale; we just need to write a function that extracts the predicted class probabilities of interest and then averages the result. The snippet of code below uses the fitted SVM from the iris example of Section~\ref{sec:class} to extract the average predicted probability of belonging to the \code{Setosa} class.
\begin{example}
pred.prob <- function(object, newdata) {  # see ?predict.svm
  pred <- predict(object, newdata, probability = TRUE)
  prob.setosa <- attr(pred, which = "probabilities")[, "setosa"]
  mean(prob.setosa)
}
\end{example}
Next, we simply pass this function via the \code{pred.fun} argument in the call to \code{partial}. The following chunk of code uses \code{partial} and \code{grid.arrange} to construct PDPs for \code{Petal.Width} and \code{Petal.Length} on the probability scale. The results are displayed in Figure~\ref{fig:partial_iris_prob}.
\begin{example}
# Display plots side by side (Figure 8)
grid.arrange(
  partial(iris.svm, pred.var = "Petal.Width", pred.fun = pred.prob, plot = TRUE),
  partial(iris.svm, pred.var = "Petal.Length", pred.fun = pred.prob, plot = TRUE),
  partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
          pred.fun = pred.prob, plot = TRUE),
  ncol = 3
)
\end{example}

@@

For regression problems, the default prediction function is essentially
\begin{example}
pred.fun <- function(object, newdata) {
  mean(predict(object, newdata), na.rm = TRUE)
}
\end{example}
This corresponds to step step 1. (c) in Algorithm~\ref{alg:pdp}. Suppose we would like ICE curves instead. To accomplish this we simply need to pass a prediction function that returns a vector of predictions, one for each observation in \code{newdata}. This is demonstarted for the Boston housing example in the code snippet below.
\begin{example}
lstat.ice <- partial(boston.rf, pred.var = "lstat",
                     pred.fun = function(object, newdata) {
                       predict(object, newdata)
                     })  # see Figure 9 (left)
\end{example}

When the function supplied to \code{pred.fun} returns multiple predictions, the data frame returned by \code{partial} includes an additional column, \code{yhat.id}, that indicates which observation a curve belongs to; in this example, there is one curve for each unique observation in \code{boston}. To further illustrate, the code below constructs ICE curves for specific quantiles of the predictions. The results, along with the traditional PDP for \code{lstat}, are displayed in Figure~\ref{fig:partial_boston_ice}.
\begin{example}
lstat.quan <- partial(boston.rf, pred.var = "lstat",
                      pred.fun = function(object, newdata) {
                        quantile(predict(object, newdata), probs = 1:9/10)
                      })  # see Figure 9 (middle)
\end{example}
\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{partial_boston_ice}
  \caption{ICE curves and PDP depicting the relationship between \code{cmedv} and \code{lstat} for the Boston housing example (the same $y$-axis limits are used for each plot). \textit{Left}: ICE curves. \textit{Middle}: Specific quantiles of ICE curves. \textit{Right}: Standard PDP.}
  \label{fig:partial_boston_ice}
\end{figure}

Using \CRANpkg{dplyr} \citep{dplyr-pkg}, it is rather straightforward to post-process ICE curves to obtain c-ICE curves; see Equation (4) on page 49 of \citet{goldstein-peeking-2015}. The code snippet below illustrates this for the Boston housing example using the predictor \code{age}.
\begin{example}
# Use partial to obtain ICE curves
pred.ice <- function(object, newdata) predict(object, newdata)
age.ice <- partial(boston.rf, pred.var = "age", pred.fun = pred.ice)

# Post-process age.ice to obtain c-ICE curves
library(dplyr)
age.ice <- age.ice %>%
  group_by(yhat.id) %>%
  mutate(yhat.centered = yhat - first(yhat))
\end{example}
Furthermore, using \pkg{ggplot2}, we can easily display ICE curves along with the PDP. This can be accomplished by using the \code{stat\_summary} function to average the ICE curves together. The code snippet below uses \pkg{ggplot2} and \code{grid.arrange} to plot ICE curves and c-ICE curves for \code{age} in the Boston housing example. The results are displayed in Figure~\ref{fig:partial_boston_ice_pdp}; compare these to Figures 3-4 in \citet{goldstein-peeking-2015}.
\begin{example}
# ICE curves
p1 <- ggplot(age.ice, aes(age, yhat)) +
  geom_line(aes(group = yhat.id), alpha = 0.2) +
  stat_summary(fun.y = mean, geom = "line", col = "red", size = 1)

# c-ICE curves
p2 <- ggplot(age.ice, aes(age, yhat.centered)) +
  geom_line(aes(group = yhat.id), alpha = 0.2) +
  stat_summary(fun.y = mean, geom = "line", col = "red", size = 1)

# Display plots side by side (Figure 10)
grid.arrange(p1, p2, ncol = 2)
\end{example}
\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{partial_boston_ice_pdp}
  \caption{ICE curves (black curves) and PDP (red curve) depicting the relationship between \code{cmedv} and \code{age} for the Boston housing example. \textit{Left}: Uncentered. \textit{Right}: Centered.}
  \label{fig:partial_boston_ice_pdp}
\end{figure}

@@

# Train an XGBoost model
set.seed(202)
boston.xgb <- train(x = data.matrix(subset(boston, select = -cmedv)),
                    y = boston$cmedv,
                    method = "xgbTree",
                    metric = "RMSE",
                    trControl = ctrl,
                    tuneGrid = xgb.grid)
\end{example}
The model with the smallest root-mean-squared error (RMSE) had \code{nrounds = 3000}, \code{max\_depth = 3}, \code{eta = 0.01}, and \code{subsample = 0.75}; the final model achieved a cross-validated RMSE and $R^2$ of 2.789 and 0.906, respectively. The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together, and then uses \code{grid.arrange} to display all three PDPs on the same page. The results are displayed in Figure~\ref{fig:boston_xgb}.
\begin{example}
# Display plots side by side (Figure 11)
grid.arrange(
  partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE),
  partial(boston.xgb, pred.var = "lstat", plot = TRUE, rug = TRUE),
  partial(boston.xgb, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE),
  ncol = 3
)
\end{example}
